% front pages for shepard book

\setcounter{page}{1}

% sub title

\thispagestyle{empty}
\begingroup
\parindent0pt
\sffamily\LARGE\selectfont\relax
\vspace*{56pt}  % 12pc - topskip - headheight - headsep - LARGE baseline
CATALOG OF TERATOGENIC AGENTS
\endgroup
\clearemptydoublepage

% title

\thispagestyle{empty}
\begingroup
\parindent0pt
{\sffamily\HUGE\bfseries\selectfont
\vspace*{34pt}  % 12pc - topskip - headheight - headsep - HUGE baseline
 CATALOG OF

 TERATOGENIC AGENTS

}


\vspace*{2pc}
\Large
Twelfth Edition

\vspace*{3pc}
{\sffamily\mediumHUGE\selectfont
Thomas H. Shepard, M.D.

}
\vspace*{1pc}
\large
Birth Defects Research Laboratory

Department of Pediatrics

School of Medicine

University of Washington

Seattle, Washington

\vspace*{1pc}

and

\vspace*{1pc}

{\sffamily\mediumHUGE\selectfont
Ronald J. Lemire, M.D.

}
\vspace*{1pc}
\large
Department of Pediatrics

School of Medicine

University of Washington

and

Children's Hospital and Regional Medical Center

Seattle, Washington

\vspace*{3pc}
{\sffamily\Large
THE JOHNS HOPKINS  UNIVERSITY  PRESS

\vspace{2pt}
Baltimore

}
\endgroup

\clearpage

% copyright


\thispagestyle{empty}
\begingroup
\parindent0pt
\def\baselinestretch{1.15}
\mediumlarge
\vspace*{56pt}  % 12pc - topskip - headheight - headsep - normal baseline
First edition (1973) sponsored by the National Institute of 
Child Health and Human Development, National Institutes of Health,
Bethesda, Maryland 20892
\vspace{1pc}

$\copyright$\ 1973, 1976, 1980, 1983, 1986, 1989, 1992, 1995, 1998, 2001, 2004, 2007
The Johns Hopkins University Press

All rights reserved.  Published 2007

Printed in the United States of America on acid-free paper

9 8 7 6 5 4 3 2 1

\vspace{1pc}

The Johns Hopkins University Press

2715 North Charles Street

Baltimore, Maryland 21218-4363

www.press.jhu.edu

\vspace{1pc}

{\it The Library of Congress has cataloged the eleventh edition of this book as
follows: }

\vspace{4pt}

Shepard, Thomas H., 1923--

\emspace Catalog of teratogenic agents / Thomas H. Shepard -- 11th ed.

\emspace \emspace p.\emspace\emspace cm.

\emspace Includes bibliographical references and indexes.

\emspace ISBN 0-8018-7953-1 (hardcover : alk. paper)

\emspace 1.~Teratogenic agents---Catalogs.

\emspace [DNLM: 1.~Teratogens---Catalogs.
                2.~Abnormalities, Drug-Induced---Catalogs.
                3.~Abnormalities, Radiation-Induced---Catalogs.
                QS~626~S547c~2004]
                I.~Lemire, Ronald J.,~1933--
                II.~Title.

\emspace QM691.S53\emspace 2004

\emspace $616'$.043---dc22 \hspace{3em} 2004004199

\vspace{2pc}

\emspace ISBN 13: 978-0-8018-8742-0 (hardcover : alk. paper)

\emspace ISBN 10: 0-8018-8742-9 (hardcover : alk. paper)


\endgroup

\clearpage

% dedication


\thispagestyle{empty}
\begingroup
\parindent0pt
\mediumlarge
\vspace*{56pt}  % 12pc - topskip - headheight - headsep - mediumlarge baseline
\def\nam#1{\par\hspace*{1pc}#1}
\raggedright
This book is dedicated with appreciation to Dr.~Shepard's past
research fellows who stimulated its production.

\vspace{3pt}

\nam{Prasanta K. Datta}
\nam{Sorrel Waxman}
\nam{Pierre Ferrier}
\nam{Charles P. Mahoney}
\nam{Ronald J. Lemire}
\nam{Ralph R. Hollingsworth}
\nam{Thomas Nelson}
\nam{Takashi Tanimura}
\nam{Alan G. Fantel}
\nam{Godfrey P. Oakley}
\nam{Gerald J. Bargman}
\nam{Theodore H. Regimbal}
\nam{Larklyn H. Fisher}
\nam{Trent D. Stephens}
\nam{Philip E. Mirkes}
\nam{Kohei Shiota}
\nam{Margo I. Van Allen}
\nam{Marlissa Campbell}
\nam{John Rogers}
\nam{Carol J. Quaife}
\nam{Jack M. FitzSimmons}
\nam{Masahiko Fujinaga}
\nam{Raj P. Kapur}
\nam{Hyoung W. Park}

\vspace{3pt}

The twelfth edition is dedicated to our wives, Gretchen and Kathy.

\endgroup

\clearemptydoublepage


\tableofcontents
\vspace*{3pc}
{\noindent\large Endpapers: Comparative Time Periods of Embryonic
and Fetal\par
\noindent Development in Humans and Experimental Animals\par}
\clearemptydoublepage



